Sperm do not evolve to collaborate in female meiotic drive.

### 2014-05-13 writeup_and_notes/male_dependent_drive.tex
In the process of overhauling the in text appendix. Finished the single-locus, sperm dep model

For comparative purposes, we present the standard drive model in equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model \citep[e.g. ][]{Prout1973,Ubeda2004} for additional results. 
In this case, assuming the driving allele is deleterious in both sexes but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$) a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. The
        parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Invasion_space}A, and where it can fix in the green
        and orange areasin Figure \ref{Invasion_space}B. When the allele is
        maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

@@

When a self-promoting driver does spread, 
	it spends much time at low frequency, 
	because the paucity of complimentary sperm compromises its ability to drive. 
However, once relatively common, it  rapidly achieves fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
This lack of a balanced polymorphism (Figure \ref{Invasion_space}) precludes the evolution of an
	allele that suppresses this form of meiotic drive in females.

%I SUBSTANTIALLY REWROTE THIS
The positive frequency dependence of a self-promoting driver can
induce bistability in its spread.
%OLD GRAHAM WORDS
 %, i.e. there is a critical frequency
%above which it can spread quickly to fixation and below which it moves towards loss (an
%unstable equilibrium frequency). In 
%the appendix, we develop a simple
%approximation to this bistability. 
When $s$ is too large to allow a rare driver to increase in frequency, but small enough to allow a common driver to fix (the inequality in EQUATION XXX),
there is  an unstable equilibrium frequency (approximated in EQUATION XXX)  
	above which it can spread quickly to fixation and below which it moves towards loss 
	(see Figure \ref{Bistab_homozyg_cost_fig} for exact results).	
%OLD GRAHAM WORDS
%This bistability occurs in the region of parameter space in Figure
%\ref{Invasion_space}. In Supplementary Figure
%\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
%show that the unstable equilibrium frequency is high across much of
%this parameter space, and so in practice the self promoting allele
%only invade in the orange space. (In the appendix we develop a simple
%approximation to the unstable equilibrium frequency
%\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
%\ref{Bistab_homozyg_cost_fig}).
This unstable equilibrium frequency is high across much of
	parameter space, likely too high to be reached by drift, 
	and therefore  the fate of self-promoting drivers  will  likely be determined
	by the more restrictive invasion criteria rather than the fixation criteria.

%OLD YANIV WORDS
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-h_s$)
further constrains the evolution of a self-promoting driver. 
In fact, with any heterozygous fitness cost, a rare self-promoting
driver cannot spread deterministically. 
However, this case also displays bistability -- 
	mathematical approximation assuming Hardy-Weinberg reveal that 
	if $s < d - 1/2+s_h(3-2d)/2$, it will fix deterministically if its 
	frequency is greater than a threshold, $f_B^*$ [Equation \eqref{thresh2}. 
	See \yb{FIGURES SXX} for exact results].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
	observed as stable polymorphisms in natural populations.

@@

\begin{figure}
	\rotatebox{90}{\includegraphics[width = 0.8
          \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig2.eps}}
\caption{Dynamics in 3 allele system
{\bf (A) } The sperm-dependent transmission probabilities of alleles through
female meiosis (female germ cells depicted as circles, sperm indicated by wiggly
tails). In each case the top and bottom entry in each cell give the
transmisson probability of the left and right allele in the female
being transmitted to the egg. In the left panel we show the transmission probabilities for
  A, B, and B'. Where B' is a driver allele which is either promotes
  itself through through female meiosis when sperm (a B$^+$ allele,
  $d'>d$) or self-avoiding through the interaction of sperm with female meiosis (a B$^-$ allele,
$d'<d$). In the right panel we depict the transmission probabilities for
  A, B, and A$^-$ allele a non-driven allele that acts to reduce the
  effects of female drive through sperm.
{\bf (B)} Frequency trajectories in the three allele system. }
 \label{Eggsperm_3_allele_cartoon}
\end{figure}

@@

%Gametogenesis and meiosis are common times for genetic conflict. 
%This is a consequence of the fact that as the set of alleles that consitute
%	an individual are in the process of parting ways, 
%	with only a subset of them being assured transmission.
%In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
%to female meiosis through the entire next life cycle of the organism. 
%It is therefore in the interest of the fertilizing sperm's genome to
%make a high fitness zygote even if that means alleles in sperm
%disrupting other copies of themselves from gaining a 
%transmission advantage through female meiosis.
  \appendix
\section*{Appendix}
In our first scenerio allele $A$ is the ancestral allele and allele
$B$ is our sperm dependent drive allele.
The sperm dependence of the drive is a form of assortative
transmission (mating), and so we have to keep track of the genotype frequencies.
Our genotype frequencies at birth are $x_{AA},~x_{AB},$ and
$x_{BB}$. These three genotypes have fitnesses $w_{AA},~w_{AB},$ and
$w_{BB}$, which are identical between males and females. Because 
fitnesses are equivalent in both sexes, genotypic frequencies
are identical in males and females, but in the following
equations they are marked with $\Mars$ and $\Venus$ respectively to
allow the reader to more easily follow the logic. 
The frequency of the three genotypes at birth in the next generation is:
\begin{equation}
x'_{BB} =\frac{1}{\bar{w} } \Big( w_{BB} x_{BB}^{\Venus} (w_{BB} x_{BB}^{\Mars} +\frac{1}{2}
  w_{AB} x_{AB}^{\Mars} ) + w_{AB} x_{AB}^{\Venus} (d w_{BB}
  x_{BB}^{\Mars} +\frac{1}{2} d
w_{AB} x_{AB}^{\Mars} ) \Big),  \label{eqn.A1}
\end{equation}
\begin{equation}
x'_{AA} =\frac{1}{\bar{w} } \Big(  w_{AA} x_{AA}^{\Venus} (w_{AA} x_{AA}^{\Mars} + \frac{1}{2}
   w_{AB} x_{AB}^{\Mars}) + w_{AB} x_{AB}^{\Venus}(\frac{1}{2} w_{AA} x_{AA}^{\Mars}+ \frac{1}{2} 
  w_{AB} x_{AB}^{\Mars}) \Big),  \label{eqn.A2}
\end{equation}
and
\begin{eqnarray}
x'_{AB} = & \frac{1}{\bar{w} } \Big( w_{AA} x_{AA}^{\Venus}  ( w_{BB}  x_{BB}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} ) +w_{BB} x_{BB}^{\Venus} ( w_{AA}  x_{AA}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} )  \nonumber \\
& + x_{AB}^{\Venus}(\frac{1}{2}  w_{AA}  x_{AA}^{\Mars}+\frac{1}{2} (1-d)
 w_{AB}  x_{AB}^{\Mars} + \frac{1}{4}  w_{AB}
 x_{AB}^{\Mars} + (1-d)  w_{BB}  x_{BB}^{\Mars} ) \Big) \label{eqn.A3}
\end{eqnarray}
where the mean fitness of the population is
\begin{equation}
\bar{w} = w_{AA} x_{AA}+w_{AB}  x_{AB} +w_{BB}  x_{BB}.
\end{equation}
We can then iterate this recursion over generations, for
example to produce the frequency trajectories shown in the inset
Figure \ref{Invasion_space} by plotting $x_{BB}+ \frac{1}{2}x_{AB}$
over time. We can also use this to assess the
invasion criteria for the allele by starting allele B at very low
frequency, and seeing if it increases across the generations.

\paragraph{Recessive fitness cost of self-promoting driver}
When the self-promoting driver has a recessive fitness cost, $s$ genotype frequencies after selection and drive equal
\begin{equation}	x'_{BB} =\frac{x_{B}^2(1-s)}{\bar{w} }  \Big( F+(1-F)(x_{B}+2x_{A} d) \Big) 	\end{equation}
\begin{equation}	x'_{AA} =\frac{1}{\bar{w} }  \Big(x_{A}^2\Big) 	\end{equation}
and
\begin{equation}	x'_{AB} =\frac{2 x_{A}x_{B} }{\bar{w} }  \Big(1-x_{B}(d-1/2)(1-F) \Big)	\end{equation}

its change in frequency across generations $\Delta f_B$ is
\begin{equation}
\frac{f_A f_B^2}{\bar{w}} \left( [1-F][ d(1-2 f_A s) - f_B s-1/2] - s  F \right)
\label{deltadriver}
\end{equation}
where $F$ is the deviation from Hardy-Weinberg expectations (HWE), and $f_A$ is the frequency of the non-driving allele. 
For analytical solutions presented above, we assume that $F$ equals zero.

We used this HWE approximation to investifgate the bistability of the
invasion criteria. Under this approximation we find that a common, recessive, self-promoting driver fixes if
 $s<(2d-1)/2$. Furthermore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, a
 recessive, self-promoting driver will deterministically fix if drift,
 mutation, or migration pressure bring its frequency above approximately
\begin{equation} 
(1-2d+4ds)/(2s(2d-1)) \label{bistab_homozy_approx}
\end{equation}
 but it will be lost if it is introduced below this frequency
 (i.e. this is an unstable equilibrium point). In Supplementary Figure \ref{Bistab_homozyg_cost_fig} we show
 this approximation condition for bistability andd how it compared to
 that obtained using the recursion. This equation
 \eqref{bistab_homozy_approx} offers a reasonable, but imperfect
 approximation 
and seems to serve as a lower bound to the
true unstable equilibrium frequency.

\paragraph{cost of driver in heterozygotes}
We explored a more complex model where the self-promoting driver has a cost, $s_h$ in heterozygotes (so $w_{AB} = 1-h_s$) in addition to the cost $s$, in homozygotes.
Model construction is similar to that above.
So long as there is any heterozygous fitness cost (i.e. $s_h > 0$) the self-promoting driver cannot invade when rare;
	however, as is the case without a heterozygous cost, this system is bistable.
Assuming no deviation from Hardy-Weinberg a self-promoting driver with a heterozygous fitness cost deterministically fixes if it's frequency exceeds some threshold, $f_B^*$.
\begin{equation}
	f_B^* = \frac{a_1+\sqrt{a_1^2+8 s_h a_2}}{2 a_2}
	\label{thresh2}
\end{equation}
where, $a_1=1-2 ds_h+4 d s-2 d-3 s_h$ and $a_2=2(s-s_h)(2d-1)$.

 range of other costs to the drive alleles other than a
simple recessive cost, for example an additive cost. We modified the recursions given in eqn
\ref{eqn.A1}-\ref{eqn.A3} to feature an additive cost $s$ for each
copy of the $B$ allele. A none recessive cost to the drive allele can
cause bistability of the drive allele. The B allele is unable to enter
the population when rare, this results from the fact that
when the allele is rare ($x_B \ll 1$) sperm carrying it rarely
encounter heterozygous eggs ($\sim x^2$) but the allele still pays a
cost in heterozygous individuals.  By contrast, when the B allele is common ($x_B
\gg 0$) encounters between heterozygous eggs and $B$ allele sperm
become more common allowing drive to occur this can allow the $B$
allele to spread when common. In Supplementary Figure
\ref{bistable_additive} we show the unstable equilibrium frequency
above which the $B$ allele can invade. This equilibrium frequency was found by a simple
parameter search.  \gc{wonder if a bit of work could find this}

>>>>

For comparative purposes, we present the standard drive model as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model \citep[e.g. ][]{Prout1973,Ubeda2004} for additional results. 
In this case, assuming the driving allele is deleterious in both sexes but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$) a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. The
        parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Invasion_space}A, and where it can fix in the green
        and orange areasin Figure \ref{Invasion_space}B. When the allele is
        maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

@@

When a self-promoting driver does spread, 
	it spends much time at low frequency, 
	because the paucity of complimentary sperm compromises its ability to drive. 
However, once relatively common, it  rapidly achieves fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
This lack of a balanced polymorphism %(Figure \ref{Invasion_space}) 
	precludes the evolution of an
	allele that suppresses this form of meiotic drive in females.

%I SUBSTANTIALLY REWROTE THIS
The positive frequency dependence of a self-promoting driver can
induce bistability in its spread.
%OLD GRAHAM WORDS
 %, i.e. there is a critical frequency
%above which it can spread quickly to fixation and below which it moves towards loss (an
%unstable equilibrium frequency). In 
%the appendix, we develop a simple
%approximation to this bistability. 
When $s$ is too large to allow a rare driver to increase in frequency, but small enough to allow a common driver to fix (the inequality in Equation \ref{ineqA}),
there is  an unstable equilibrium frequency, $f^*_{B\text{ recessive}}$  
	above which it can spread quickly to fixation and below which it moves towards loss 
	($f^*_{B\text{ recessive}}$ is approximated in Equation \ref{bistab_homozy_approx}.  
	We compare this approximation to the exact result in Figure \ref{Bistab_homozyg_cost_fig}).	
%OLD GRAHAM WORDS
%This bistability occurs in the region of parameter space in Figure
%\ref{Invasion_space}. In Supplementary Figure
%\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
%show that the unstable equilibrium frequency is high across much of
%this parameter space, and so in practice the self promoting allele
%only invade in the orange space. (In the appendix we develop a simple
%approximation to the unstable equilibrium frequency
%\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
%\ref{Bistab_homozyg_cost_fig}).
This unstable equilibrium frequency is high across much of
	parameter space, likely too high to be reached by drift, 
	and therefore  the fate of self-promoting drivers  will  likely be determined
	by the more restrictive invasion criteria rather than the fixation criteria.

%OLD YANIV WORDS
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-s_h$)
further constrains the evolution of a self-promoting driver. 
In fact, with any heterozygous fitness cost, a rare self-promoting
driver cannot spread deterministically. 
However, this case also displays bistability -- 
	mathematical approximation assuming Hardy-Weinberg reveal that 
	if $s < d - 1/2+s_h(3-2d)/2$, it will fix deterministically if its 
	frequency is greater than a threshold, $f_B^*$ [Equation \eqref{thresh2}. 
	See \yb{FIGURES SXX} for exact results].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
	observed as stable polymorphisms in natural populations.

@@

\begin{figure}
	\rotatebox{90}{\includegraphics[width = 0.8
          \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig2.eps}}
\caption{Dynamics in 3 allele system
{\bf (A) } The sperm-dependent transmission probabilities of alleles through
female meiosis (female germ cells depicted as circles, sperm indicated by wiggly
tails). In each case the top and bottom entry in each cell give the
transmission probability of the left and right allele in the female
being transmitted to the egg. In the left panel we show the transmission probabilities for
  A, B, and B'. Where B' is a driver allele which is either promotes
  itself through through female meiosis when sperm (a B$^+$ allele,
  $d'>d$) or self-avoiding through the interaction of sperm with female meiosis (a B$^-$ allele,
$d'<d$). In the right panel we depict the transmission probabilities for
  A, B, and A$^-$ allele a non-driven allele that acts to reduce the
  effects of female drive through sperm.
{\bf (B)} Frequency trajectories in the three allele system. }
 \label{Eggsperm_3_allele_cartoon}
\end{figure}

@@

%Gametogenesis and meiosis are common times for genetic conflict. 
%This is a consequence of the fact that as the set of alleles that consitute
%	an individual are in the process of parting ways, 
%	with only a subset of them being assured transmission.
%In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
%to female meiosis through the entire next life cycle of the organism. 
%It is therefore in the interest of the fertilizing sperm's genome to
%make a high fitness zygote even if that means alleles in sperm
%disrupting other copies of themselves from gaining a 
%transmission advantage through female meiosis.
  \appendix

\section*{Appendix}
Below, we highlight some of our derivations and analytical results. 
The fitness of genotype, $i$ is $w_g$ and its frequency before drive, after drive, and after drive and selection are $f_g$, $f_g'$ and $f_g''$, respectively, and population mean fitness,  $\bar{w}$ equals  $\Sigma f_g'w_g$. Genotypic fitnesses are $w_{AA} = 1$, $w_{AB} = 1-s_h$, and $w_{BB}=1-s$, and are sex-independent. 
All models include a balletic locus with non-driving and driving alleles in frequencies $f_A$ and $f_B$, respectively, while more complex models include more loci and/or alleles.  
Because many of our analyses assume Hardy Weinberg Equilibrium (HWE), we note which analytical results are approximations. 
We verify these approximations with exact numerical iterations in supplementary figures and scripts, and detail the derivation and details of these results in our mathematical supplement.

\subsection*{Model 1. Traditional driver}
In the standard drive model meiosis in males is fair such that $A/B$ heterozygotes contribute $A$ and $B$ alleles with equal probabilities; however, $A/B$ females transmit the $B$ allele with probability $d>1/2$. 
After drive and random mating genotype frequencies are 
	$f_{AA}'=f_A (f_{AA }+ f_{AB} (1 - d))$, 
	$f_{AB}'=f_A (f_{BB} + f_{ABd}) + f_B (f_{AA} + f_{AB} (1 - d))$,
	and $f_{BB}= f_B (f_{AB} d + f_{BB})$. 
%Assuming fitnesses $w_{AA}=1$, $w_{AB}-1 s_h$ and $w_{BB}=1-s$, 
Frequencies of genotype, $i$, after drive and selection are 
	\begin{equation} f_i''= f_i'w_i/\bar{w}\end{equation}
	where $w_i$ is genotypic fitness, and mean fitness, $\bar{w} = \Sigma f_i'w_i$. 
	Mathematical approximations assuming HWE show that a rare driver will spread when $s_h <(2d-1)/(1+2d)$, and will fix when $s< d -1/2 + 3 s_h /2- d s_h$. This later inequality reduces to $s<(2d-1)/2$ when the cost of drive is fully recessive.

\subsection*{Model 2. Single locus, sperm-dependent drive.}
Our single-locus model of sperm dependent drive resembles the traditional driver with the caveat that the $B$ allele drives in heterozygous females only when fertilized by $B$ bearing sperm. 
Therefore, genotype frequencies after drive are 
	$f_{AA}'=f_A^2$, 
	$f_{AB}'=f_A f_B+f_B (f_{AA} + f_{AB}(1 - d)) $,
	and $f_{BB}'= f_B (f_{AB} d + f_{BB} )$. 
As above, exact frequencies after drive, random mating and selection are $f_i''= f_i'w_i/\bar{w}$. 
We iterate this recursion over generations, to produce the frequency trajectories shown in the inset
Figure \ref{Invasion_space} by plotting $x_{BB}+ \frac{1}{2}x_{AB}$
over time, and to assess the analytical approximations, below.

% the $B$ allele In our first scenerio allele $A$ is the ancestral allele and allele
%$B$ is our sperm dependent drive allele.
%The sperm dependence of the drive is a form of assortative
%transmission (mating), and so we have to keep track of the genotype frequencies.
%Our genotype frequencies at birth are $x_{AA},~x_{AB},$ and
%$x_{BB}$. These three genotypes have fitnesses $w_{AA},~w_{AB},$ and
%$w_{BB}$, which are identical between males and females. Because 
%fitnesses are equivalent in both sexes, genotypic frequencies
%are identical in males and females, but in the following
%equations they are marked with $\Mars$ and $\Venus$ respectively to
%allow the reader to more easily follow the logic. 
%The frequency of the three genotypes at birth in the next generation is:
%\begin{equation}
%x'_{BB} =\frac{1}{\bar{w} } \Big( w_{BB} x_{BB}^{\Venus} (w_{BB} x_{BB}^{\Mars} +\frac{1}{2}
%  w_{AB} x_{AB}^{\Mars} ) + w_{AB} x_{AB}^{\Venus} (d w_{BB}
%  x_{BB}^{\Mars} +\frac{1}{2} d
%w_{AB} x_{AB}^{\Mars} ) \Big),  \label{eqn.A1}
%\end{equation}
%\begin{equation}
%x'_{AA} =\frac{1}{\bar{w} } \Big(  w_{AA} x_{AA}^{\Venus} (w_{AA} x_{AA}^{\Mars} + \frac{1}{2}
%   w_{AB} x_{AB}^{\Mars}) + w_{AB} x_{AB}^{\Venus}(\frac{1}{2} w_{AA} x_{AA}^{\Mars}+ \frac{1}{2} 
%  w_{AB} x_{AB}^{\Mars}) \Big),  \label{eqn.A2}
%\end{equation}
%and
%\begin{eqnarray}
%x'_{AB} = & \frac{1}{\bar{w} } \Big( w_{AA} x_{AA}^{\Venus}  ( w_{BB}  x_{BB}^{\Mars} + \frac{1}{2}
%  w_{AB}  x_{AB}^{\Mars} ) +w_{BB} x_{BB}^{\Venus} ( w_{AA}  x_{AA}^{\Mars} + \frac{1}{2}
%  w_{AB}  x_{AB}^{\Mars} )  \nonumber \\
%& + x_{AB}^{\Venus}(\frac{1}{2}  w_{AA}  x_{AA}^{\Mars}+\frac{1}{2} (1-d)
% w_{AB}  x_{AB}^{\Mars} + \frac{1}{4}  w_{AB}
 %x_{AB}^{\Mars} + (1-d)  w_{BB}  x_{BB}^{\Mars} ) \Big) \label{eqn.A3}
%\end{eqnarray}
%where the mean fitness of the population is
%\begin{equation}
%\bar{w} = w_{AA} x_{AA}+w_{AB}  x_{AB} +w_{BB}  x_{BB}.
%\end{equation}
%We can then iterate this recursion over generations, for
%example to produce the frequency trajectories shown in the inset
%Figure \ref{Invasion_space} by plotting $x_{BB}+ \frac{1}{2}x_{AB}$
%over time. We can also use this to assess the
%invasion criteria for the allele by starting allele B at very low
%frequency, and seeing if it increases across the generations.

\paragraph{Recessive fitness cost of self-promoting driver:} %When the self-promoting driver has a recessive fitness cost, $s$ genotype frequencies after selection and drive equal
%\begin{equation}	x'_{BB} =\frac{x_{B}^2(1-s)}{\bar{w} }  \Big( F+(1-F)(x_{B}+2x_{A} d) \Big) 	\end{equation}
%\begin{equation}	x'_{AA} =\frac{1}{\bar{w} }  \Big(x_{A}^2\Big) 	\end{equation}
%and
%\begin{equation}	x'_{AB} =\frac{2 x_{A}x_{B} }{\bar{w} }  \Big(1-x_{B}(d-1/2)(1-F) \Big)	%\end{equation}
When fully recessive, the change in frequency of the self-promoting driver across generations $\Delta f_B$ is exactly 
\begin{equation}
f_A f_B^2 \left( [1-F][ d(1-2 f_A s) - f_B s-1/2] - s  F \right)/\bar{w}
\label{deltadriver}
\end{equation}
where $F$ is the deviation from Hardy-Weinberg expectations (HWE). 
%For analytical solutions presented above, we assume that $F$ equals zero. 
Assuming HWE a common, recessive, self-promoting driver invades if $(2 d - 1)/(4 d)$, and fixes if
 $s<(2d-1)/2$. 
 Therefore, when \begin{equation}(2 d - 1)/(4 d)<s<(2d-1)/2\label{ineqA}\end{equation} a
 	recessive, self-promoting driver will deterministically fix if drift,
 	mutation, or migration pressure bring its frequency above approximately
\begin{equation} 
f^*_{B\text{ recessive}} = (1-2d+4ds)/(2s(2d-1)) \label{bistab_homozy_approx}
\end{equation}
 but it will be lost if it is introduced below this frequency. 
 %(i.e. this is an unstable equilibrium point). 
% In Supplementary Figure \ref{Bistab_homozyg_cost_fig} we show
% this approximation condition for bistability andd how it compared to
% that obtained using the recursion. 
Compared to exact results (Supplementary Figure \ref{Bistab_homozyg_cost_fig}), Equation
 \eqref{bistab_homozy_approx} offers a reasonable, but imperfect approximation. 
%s_h and seems to serve as a lower bound to the true unstable equilibrium frequency.

\paragraph{Cost of driver in heterozygotes:} %We explored a more complex model where the self-promoting driver has a cost, $s_h$ in heterozygotes (so $w_{AB} = 1-s_h$) in addition to the cost $s$, in homozygotes.
When the fitness of drive heterozygotes is compromised ($s_h>0$), a rare, self promoting driver cannot invade when rare.
This results from the fact that
when the allele is rare ($f_B \ll 1$) sperm carrying it rarely
encounter heterozygous eggs ($\sim f_B^2$) but the allele still pays a
cost in heterozygous individuals. 
%Model construction is similar to that above.
%So long as there is any heterozygous fitness cost (i.e. $s_h > 0$) the self-promoting driver cannot invade when rare;
However, this system too, is bistable -- as the driver becomes more frequency it is more often fertilized by a driving sperm and therefore drives more effectively.
%Assuming no deviation from Hardy-Weinberg a self-promoting driver with a heterozygous fitness cost 
This allele deterministically fixes if its frequency exceeds some threshold, $f_B^*$.
Assuming HWE, this threshold equals 
\begin{equation}
	f_B^* = \frac{a_1+\sqrt{a_1^2+8 s_h a_2}}{2 a_2}
	\label{thresh2}
\end{equation}
where, $a_1=1-2 ds_h+4 d s-2 d-3 s_h$ and $a_2=2(s-s_h)(2d-1)$.
\yb{We compare this analytical approximation to exact results obtained by a simple parameter search in Supplementary Figure \ref{bistable_additive}. }
 %range of other costs to the drive alleles other than a
%simple recessive cost, for example an additive cost. We modified the recursions given in eqn
%\ref{eqn.A1}-\ref{eqn.A3} to feature an additive cost $s$ for each
%copy of the $B$ allele. A none recessive cost to the drive allele can
%cause bistability of the drive allele. The B allele is unable to enter
%the population when rare, this results from the fact that
%when the allele is rare ($x_B \ll 1$) sperm carrying it rarely
%encounter heterozygous eggs ($\sim x^2$) but the allele still pays a
%cost in heterozygous individuals.  By contrast, when the B allele is common ($x_B
%\gg 0$) encounters between heterozygous eggs and $B$ allele sperm
%become more common allowing drive to occur this can allow the $B$
%allele to spread when common. In Supplementary Figure
%\ref{bistable_additive} we show the unstable equilibrium frequency
%above which the $B$ allele can invade. This equilibrium frequency was found by a simple
%parameter search.  \gc{wonder if a bit of work could find this}
